---
title: "Staffed by Size, Not by Need"
subtitle: "Pupil/teacher ratios in Kansas public schools, 2023-24 · DS 3100 Analysis Report 1"
author: "Aneesh Vasamreddy"
output:
  html_document:
    self_contained: true
    theme: flatly
    toc: true
    toc_float:
      collapsed: true
    code_folding: hide
---

<style>
.main-container { max-width: 900px; }
.html-widget, .plotly { max-width: 100% !important; }
table { display: block; overflow-x: auto; }
.caption, .figcap { color: #555; font-size: 0.92em; margin-bottom: 1.2em; }
.answer { background: #f4f6f8; border-left: 4px solid #B2182B;
          padding: 0.8em 1em; margin: 1em 0; }
@media (max-width: 600px) { body { font-size: 15px; } h1.title { font-size: 1.7em; } }
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      out.width = "100%")
library(tidyverse)
library(scales)
library(plotly)
library(crosstalk)
library(DT)
```

```{r data-and-numbers}
# Same reading, wrangling and numbers as the PDF report
schools_raw <- read_csv("data/kansas_public_schools_2023-24.csv",
                        skip = 6, n_max = 1348)
# Short column names in place of NCES's long ones
names(schools_raw) <- c("school", "state", "id", "district", "county",
                        "city", "level", "locale", "charter", "virtual",
                        "students", "frl", "teachers", "ratio",
                        "male", "female")

schools <- schools_raw |>
  # NCES symbols for missing / not applicable / low quality become NA,
  # so the three columns can be numbers
  mutate(across(c(frl, teachers, ratio),
                ~ .x |> na_if("\u2020") |> na_if("\u2013") |>
                  na_if("\u2021") |>
                  as.numeric()),
         state = str_to_title(state))

# Virtual schools have no classrooms, so their ratios mean something else
in_person <- schools |> filter(virtual == "NOTVIRTUAL")

# Missing values in the variables the report uses, before dropping
missing_counts <- in_person |>
  summarize(students = sum(is.na(students)), frl = sum(is.na(frl)),
            ratio = sum(is.na(ratio)))

analysis <- in_person |>
  mutate(frl_share = frl / students,
         size_band = case_when(students < 200 ~ "Under 200",
                               students < 600 ~ "200 to 599",
                               TRUE ~ "600 or more") |>
           fct_relevel("Under 200", "200 to 599", "600 or more")) |>
  drop_na(frl_share, ratio)
r_frl     <- cor(analysis$frl_share, analysis$ratio)
low_need  <- median(analysis$ratio[analysis$frl_share < 0.5])
high_need <- median(analysis$ratio[analysis$frl_share >= 0.5])
r_size     <- cor(analysis$students, analysis$ratio)
r_log_size <- cor(log10(analysis$students), analysis$ratio)
small <- median(analysis$ratio[analysis$size_band == "Under 200"])
large <- median(analysis$ratio[analysis$size_band == "600 or more"])
by_band <- analysis |>
  group_by(size_band) |>
  summarize(schools = n(),
            median_frl = median(frl_share),
            median_ratio = median(ratio),
            r = cor(frl_share, ratio))
r_size_frl <- cor(analysis$students, analysis$frl_share)
```

<div class="answer">
**The short answer.** Across `r nrow(analysis)` in-person Kansas public
schools, the share of students on free or reduced-price lunch says almost
nothing about how many students each teacher has (correlation
`r round(r_frl, 2)`), but school size does: `r small` students per teacher in
schools under 200 students against `r large` in schools of 600 or more.
Hover over any point to see the school, or filter by size below to check it.
</div>

## Background and question

Teachers are the largest cost in a school budget, and schools that serve
poorer communities are often assumed to run larger classes. Kansas reports,
for every public school, both the share of students eligible for free or
reduced-price lunch (a standard measure of student poverty) and the number
of students per teacher. Putting the two side by side shows whether the
state's higher-need schools are more thinly staffed, and if not, what does go
with staffing.

**Among Kansas's in-person public schools in 2023-24, how does the
pupil/teacher ratio (students per FTE teacher) relate to the share of
students eligible for free or reduced-price lunch (0 to 100%), and is
school enrollment (students) more closely related to the ratio than that
share is, including among schools of similar size?**

**Data.** NCES Common Core of Data, Public Elementary/Secondary School
Universe Survey, 2023-24 v.1a, via the
[ELSI Table Generator](https://nces.ed.gov/ccd/elsi/tableGenerator.aspx).
The `r sum(schools_raw$virtual == "FULLVIRTUAL")` full-virtual schools and the
`r nrow(in_person) - nrow(analysis)` schools missing a lunch count or ratio are
left out. Expand the *Code* buttons to see every step.

## Figure 1: Need and staffing

```{r fig1-web}
hover_text <- ~ paste0("<b>", str_to_title(school), "</b><br>",
                       str_to_title(district), " district<br>",
                       "Enrollment: ", comma(students), "<br>",
                       "Lunch share: ", percent(frl_share, 1), "<br>",
                       "Students per teacher: ", ratio)
fit1 <- lm(ratio ~ frl_share, data = analysis)
line1 <- tibble(frl_share = c(min(analysis$frl_share),
                              max(analysis$frl_share))) |>
  mutate(ratio = predict(fit1, newdata = pick(everything())))

plot_ly(analysis, x = ~frl_share, y = ~ratio, type = "scatter",
        mode = "markers", text = hover_text, hoverinfo = "text",
        marker = list(color = "rgba(40,40,40,0.35)", size = 6),
        name = "School") |>
  add_lines(data = line1, x = ~frl_share, y = ~ratio, inherit = FALSE,
            line = list(color = "#B2182B", width = 3),
            name = "Least-squares fit", hoverinfo = "skip") |>
  layout(xaxis = list(title = "Students eligible for free or reduced-price lunch (%)",
                      tickformat = ".0%"),
         yaxis = list(title = "Students per FTE teacher", rangemode = "tozero"),
         showlegend = FALSE) |>
  config(displayModeBar = FALSE)
```

<p class="figcap">**Figure 1.** Pupil/teacher ratio against the share of students eligible for free or reduced-price lunch, for Kansas's in-person public schools in 2023-24. One point is one school; the line is a least-squares fit. The line is nearly flat, dipping slightly as need rises.</p>

**Interpretation of Figure 1.** Figure 1 plots the two variables the question
names for all `r nrow(analysis)` schools. The feature to look at is the
fitted line, which is nearly flat and, if anything, tilts down: schools where
at least half the students qualify for free or reduced-price lunch have a
median of `r high_need` students per teacher, against `r low_need` for schools
below half, and the correlation is `r round(r_frl, 2)`. For the question, this
means Kansas's higher-need schools are not more thinly staffed. The figure
cannot show whether something that travels with lunch share, such as school
size or rural location, is hiding or producing that pattern.

## Figure 2: Size and staffing

```{r fig2-web}
fit2 <- lm(ratio ~ log10(students), data = analysis)
line2 <- tibble(students = 10^seq(log10(min(analysis$students)),
                                  log10(max(analysis$students)),
                                  length.out = 50)) |>
  mutate(ratio = predict(fit2, newdata = pick(everything())))

plot_ly(analysis, x = ~students, y = ~ratio, type = "scatter",
        mode = "markers", text = hover_text, hoverinfo = "text",
        marker = list(color = "rgba(40,40,40,0.35)", size = 6)) |>
  add_lines(data = line2, x = ~students, y = ~ratio, inherit = FALSE,
            line = list(color = "#B2182B", width = 3), hoverinfo = "skip") |>
  layout(xaxis = list(title = "Students enrolled (log scale)", type = "log"),
         yaxis = list(title = "Students per FTE teacher", rangemode = "tozero"),
         showlegend = FALSE) |>
  config(displayModeBar = FALSE)
```

<p class="figcap">**Figure 2.** Pupil/teacher ratio against total enrollment (log scale), one point per school, with a least-squares fit on log enrollment. Larger schools have more students per teacher.</p>

**Interpretation of Figure 2.** Figure 2 keeps the same vertical axis as
Figure 1 but swaps lunch share for enrollment, which runs from
`r min(analysis$students)` to `r comma(max(analysis$students))` students and
so is drawn on a log scale. The feature to look at is the upward slope:
schools under 200 students have a median of `r small` students per teacher,
against `r large` for schools of 600 or more, and the correlation of the ratio
with log enrollment is `r round(r_log_size, 2)`, compared with
`r round(r_frl, 2)` for lunch share. For the question, enrollment goes with
staffing far more closely than need does. The figure cannot say why; small
schools may need a minimum number of teachers however few students they have,
but the data do not show that directly.

## Figure 3: Need and staffing among schools of similar size

```{r fig3-web, fig.height = 4}
fig3 <- ggplot(analysis, aes(x = frl_share, y = ratio, text = school)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(aes(group = size_band), method = "lm", se = FALSE,
              color = "#B2182B") +
  facet_wrap(~ size_band) +
  scale_x_continuous(labels = percent, breaks = c(0.25, 0.75)) +
  labs(x = "Students eligible for free or reduced-price lunch (%)",
       y = "Students per FTE teacher") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.spacing = unit(1, "lines"))
ggplotly(fig3, tooltip = c("text", "x", "y")) |>
  config(displayModeBar = FALSE)
```

<p class="figcap">**Figure 3.** Figure 1 redrawn for three enrollment bands, on a common vertical scale. One point is one school; the line is a least-squares fit within each band. The bands sit at different heights while the line inside each stays close to flat.</p>

```{r table3-web}
knitr::kable(by_band, digits = 2,
             col.names = c("Enrollment", "Schools", "Median lunch share",
                           "Median students per teacher",
                           "Correlation with lunch share"))
```

**Interpretation of Figure 3.** Figure 3 draws Figure 1 again for three
enrollment bands, so that each school is compared only with schools of
similar size. The feature to look at is that the panels sit at different
heights, with medians of `r by_band$median_ratio[1]`,
`r by_band$median_ratio[2]` and `r by_band$median_ratio[3]` students per
teacher, while the line inside each stays close to flat, with correlations of
`r round(by_band$r[1], 2)`, `r round(by_band$r[2], 2)` and
`r round(by_band$r[3], 2)` (table below). For the question, size moves a whole
panel up or down, but within a size band lunch share still says little about
staffing. The figure cannot establish that size causes the ratio: the bands
are cut-offs I chose, and being small also goes with being rural.

## Check it yourself

Pick a size band, or narrow the lunch-share range, and the plot and table
update together. With a band selected, the table's median and the slope of
the cloud let you check the claims above, for instance that small schools
sit lower whatever their lunch share. Search the table for any school by
name.

```{r explore}
explore <- analysis |>
  transmute(School = str_to_title(school),
            District = str_to_title(district),
            `Size band` = size_band,
            Enrollment = students,
            `Lunch share (%)` = round(100 * frl_share, 1),
            `Students per teacher` = ratio)
shared <- SharedData$new(explore)

bscols(widths = c(12, 12),
  bscols(widths = c(6, 6),
    filter_checkbox("band", "Enrollment band", shared, ~`Size band`,
                    inline = TRUE),
    filter_slider("frl", "Lunch share (%)", shared, ~`Lunch share (%)`,
                  step = 1, width = "100%")),
  plot_ly(shared, x = ~`Lunch share (%)`, y = ~`Students per teacher`,
          color = ~`Size band`, colors = c("#67A9CF", "#2166AC", "#B2182B"),
          type = "scatter", mode = "markers",
          text = ~paste0("<b>", School, "</b><br>Enrollment: ", Enrollment),
          hoverinfo = "text+x+y", marker = list(size = 6, opacity = 0.6)) |>
    layout(xaxis = list(title = "Students eligible for free or reduced-price lunch (%)"),
           yaxis = list(title = "Students per FTE teacher"),
           legend = list(orientation = "h", y = -0.25)) |>
    config(displayModeBar = FALSE),
  datatable(shared, rownames = FALSE, width = "100%",
            options = list(pageLength = 8, scrollX = TRUE,
                           order = list(list(5, "desc"))))
)
```

## What we learned

**Summary.** Kansas's higher-need schools are not more thinly staffed. The
pupil/teacher ratio barely moves with lunch share (Figure 1, correlation
`r round(r_frl, 2)`), but it rises with enrollment, from `r small` students
per teacher in schools under 200 students to `r large` in schools of 600 or
more (Figure 2). Comparing schools of the same size does not change this
(Figure 3), and since lunch share and enrollment are barely correlated
(`r round(r_size_frl, 2)`), size is not hiding a need effect. In 2023-24,
Kansas schools were staffed in line with their size, not their students'
need.

---

*Source: U.S. Department of Education, National Center for Education
Statistics, Common Core of Data 2023-24 v.1a. Full report with Data,
wrangling, Implications, Limitations and Future directions in the PDF
submitted on Brightspace. Built with R Markdown, plotly, crosstalk and DT.*
